---
title: "Assignment 5: Data Visualization"
author: "Justin Maynard"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---
```{r}
#Clear environment
#rm(list = ls())
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)
here()
NTL_LTER_chemistry <- read.csv(file = "Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
NEON_NIWO_litter <- read.csv(file = "Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv")

#2 
class(NTL_LTER_chemistry$sampledate)
class(NEON_NIWO_litter$collectDate)

NTL_LTER_chemistry$sampledate <- ymd(NTL_LTER_chemistry$sampledate )
NEON_NIWO_litter$collectDate <- ymd(NEON_NIWO_litter$collectDate)
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
my_theme <-
  theme(
      plot.background = element_rect(fill = "#f5f0d9", color = NA),
      panel.background = element_rect(fill = "#f5f0d9", color = NA),
      legend.background = element_rect(fill = "#f5f0d9", color = NA),
      panel.grid.major =  element_line(color = "#d0d0d0", linewidth = .2),
      legend.position="top")
      
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
q4 <- ggplot(NTL_LTER_chemistry,
             aes(
               x = tp_ug,
               y = po4,
               color = lakename
             )) +
  geom_point() + 
  geom_smooth(method = lm, color = "black") + 
  ylim(c(0,50)) + my_theme
q4
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 
library(cowplot)

q5a <- ggplot(NTL_LTER_chemistry, 
              aes(
                x = month.abb[month],
                y = temperature_C,
                color = lakename
              )) + 
  geom_boxplot(fill = "grey90") + 
  my_theme + 
  scale_x_discrete(limits = month.abb)
q5a

q5b <- ggplot(NTL_LTER_chemistry, 
              aes(
                x = month.abb[month],
                y = tp_ug,
                color = lakename
              )) + 
  geom_boxplot(fill = "grey90") + my_theme + 
  scale_x_discrete(limits = month.abb)
q5b

q5c <- ggplot(NTL_LTER_chemistry, 
              aes(
                x = month.abb[month],
                y = tn_ug,
                color = lakename
              )) + 
  geom_boxplot(fill = "grey90") + my_theme + 
  scale_x_discrete(limits = month.abb)
q5c

q5_final <- plot_grid(q5a + theme(legend.position="top"),
          q5b + theme(legend.position = "none"),
          q5c + theme(legend.position = "none"),
          nrow = 3,
          align = 'h',
          rel_heights = c(1.25, 1)) + my_theme
q5_final

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: The temperature increases during the summer months, while the tp_ug slighly increases for Peter Lake in the late summer. The tn_ug also slightly increases for Peter lake in the late summer.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
q6 <- ggplot(subset(NEON_NIWO_litter, functionalGroup == "Needles"),
       aes(
         x = collectDate,
         y = dryMass,
         color = nlcdClass
       )) +
  geom_point() + my_theme
q6

#7
q7 <- ggplot(subset(NEON_NIWO_litter, functionalGroup == "Needles"),
       aes(
         x = collectDate,
         y = dryMass,
         color = nlcdClass
       )) +
  geom_point() +
  facet_wrap(vars(nlcdClass), nrow = 3) + my_theme
q7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Plot 7 is more effective as it sepeates out the nlcd class type, allowing one to visualize the change in dry mass for each class by date. Plot 6 is too busy to try and determine change in dry mass by class by year.